---
title: "Quantile inference and study planning with elquantile"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
    mathjax: null
    pandoc_args: ["--mathml"]
bibliography: references.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Quantile inference and study planning with elquantile}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", error = FALSE, cache = FALSE,
  fig.width = 6.2, fig.height = 3.8, dpi = 96, out.width = "100%"
)
library(elquantile)
```

This tutorial follows an analysis from a one-sample quantile question to
independent-group comparisons and study planning. It also explains how to
compare Bartlett corrections and inspect numerical diagnostics. All displayed
results are calculated from the accompanying code; the small simulations are
examples, not a comparison of the methods' statistical performance.

## Choose the scientific target

Let $q_\tau$ be a population quantile at probability $\tau$, such as the
median at $\tau=0.5$. Under the continuous-population assumptions used here,
$P(X\leq q_\tau)=\tau$. The EL constructions use this identity without
specifying an entire parametric distribution for inference.

| Design | Target | Main function |
|:---|:---|:---|
| One sample | Population quantile | `elq_one_sample()` |
| Paired observations | Quantile of within-pair differences $X-Y$ | `elq_two_sample(paired = TRUE)` |
| Two independent samples | Difference of marginal quantiles $q_{\tau,X}-q_{\tau,Y}$ | `elq_two_sample()` |
| Independent groups | Equality of their quantiles at one specified $\tau$ | `elq_k_sample()` or `elq_anova()` |

The quantile of a difference is generally not the difference of quantiles.
Specify the target before choosing a method. Unequal sample sizes and group
variances are allowed in the independent-group designs.

The intended setting has independent observations, or independent pairs,
unique target quantiles, and positive densities in their neighborhoods.
Smoothing methods require additional kernel, bandwidth, and smoothness
conditions. Clustered observations, dependent repeated measurements beyond
independent pairs, censored responses, and discrete-population quantile
inference require separate methods.

The current implementation requires finite numeric vectors and **rejects
ties**. This is a conservative implementation restriction, not a universal
requirement of empirical likelihood. Many EL criteria are computable with
repeated values, but interpolation, bandwidth choice, and calibration need
method-specific treatment. Do not add arbitrary jitter simply to bypass the
check. Missing observations are not dropped automatically.

## Start with one sample

Generate a reproducible sample from an exponential distribution. Its
population median is $\log(2)$; the distribution is used here to make the
example's target known, not as an assumption supplied to the inference engine.

```{r one-sample}
set.seed(2701)
x <- rexp(80)
tau <- 0.5
q0 <- log(2)

one <- elq_one_sample(
  x, q0 = q0, tau = tau, method = "chen_hall", conf.int = TRUE
)
one
```

The output reports the empirical quantile estimate, statistic, p-value, and a
two-sided confidence interval. A small p-value is evidence against the
specified quantile. The printed estimate uses R's type-1 empirical quantile;
it need not equal the minimizer of a smoothed likelihood criterion.

All scalar method statistics use the $-2\log R$ scale, or the stated
likelihood-ratio analogue, with first-order $\chi^2_1$ calibration.

| `method` | Construction |
|:---|:---|
| `"el"` | Ordinary indicator-based EL |
| `"adimari"` | Midpoint interpolation in the explicit Bernoulli KL criterion |
| `"chen_hall"` | Kernel-smoothed estimating equations and an EL dual solve |
| `"chen_hall_bartlett"` | Chen-Hall statistic divided by a Bartlett scale |
| `"zhou_jing"` | A signed-kernel distribution estimate in the Bernoulli KL criterion |
| `"ael"` | EL with an added pseudo estimating value |

Adimari's interpolation [@adimari1998], Chen-Hall smoothing
[@chenhall1993], and Zhou-Jing substitution [@zhoujing2003] are distinct
constructions. The AEL option adapts the estimating-equation adjustment of
@chen2008 to quantiles; it is different from Zhou-Jing's use of "adjusted."
An AEL statistic is finite even when ordinary EL has no positive-likelihood
solution.

### Inspect a likelihood curve

A likelihood curve shows which candidate quantiles are compatible with the
data under the selected criterion. The horizontal line is the first-order
95% confidence-set cutoff.

```{r likelihood-curve, fig.cap="Adimari criterion for the exponential sample. Candidate quantiles below the dashed chi-square cutoff form the accepted region."}
profile <- elq_profile(
  x, theta = seq(0.2, 1.5, length.out = 180),
  tau = tau, method = "adimari"
)
plot(profile, conf.level = 0.95)
```

Confidence sets are returned as **closed hulls**. Where the method permits
disconnected sets, inspect the retained components rather than interpreting
every point in the hull as accepted by the test:

```{r confidence-components}
zj_ci <- elq_confint(x, tau = tau, method = "zhou_jing")
zj_ci
attr(zj_ci, "components")
```

The Zhou-Jing kernel is signed; its integrated function can be nonmonotone.
The implementation retains that feature and assigns an infinite statistic
outside its valid open probability branch. AEL intervals can be unbounded.

## Compare the two Bartlett corrections

Chen-Hall uses
\[
g_i(\theta)=G\!\left(\frac{\theta-X_i}{h}\right)-\tau.
\]
For the empirical correction, the package calculates
\[
\widehat\mu_r=\frac1n\sum_{i=1}^n
\left\{G\!\left(\frac{\widehat q_\tau-X_i}{h}\right)-\tau\right\}^r,
\qquad
\widehat\beta=
\frac{\widehat\mu_4}{2\widehat\mu_2^2}
-\frac{\widehat\mu_3^2}{3\widehat\mu_2^3}.
\]
These are moments of the estimating values about zero. The quantile
$\widehat q_\tau$ is the ordinary type-1 sample quantile.

The simpler correction uses the population factor's limit as $h\to0$:
\[
\beta_0=\frac{1-\tau+\tau^2}{6\tau(1-\tau)}.
\]
At the median, $\beta_0=0.5$. This constant is a **correction factor**,
not the test statistic. Both variants retain the sample-dependent smoothed
statistic and calculate $W_{\mathrm{BC}}=W_h/(1+\beta/n)$.

Use the same kernel and explicit bandwidth to isolate the effect of the
correction:

```{r bartlett-comparison}
h <- elq_bandwidth(x, method = "chen_hall_bartlett")
b_plugin <- elq_bartlett(x, tau = tau, bandwidth = h, type = "plugin")
b_limit <- elq_bartlett(x, tau = tau, type = "limit")

w <- c(
  uncorrected = elq_statistic(
    x, theta = q0, tau = tau, method = "chen_hall", bandwidth = h
  ),
  plugin = elq_statistic(
    x, theta = q0, tau = tau, method = "chen_hall_bartlett",
    bandwidth = h, bartlett = "plugin"
  ),
  limit = elq_statistic(
    x, theta = q0, tau = tau, method = "chen_hall_bartlett",
    bandwidth = h, bartlett = "limit"
  )
)
knitr::kable(
  data.frame(
    version = names(w), factor = c(0, b_plugin, b_limit),
    statistic = unname(w), p.value = pchisq(w, df = 1, lower.tail = FALSE)
  ),
  row.names = FALSE, digits = 4
)
```

The default kernel is biweight. Default bandwidths use
$\operatorname{IQR}(x)/(2\Phi^{-1}(0.75))$ multiplied by $n^{-1/2}$
for uncorrected Chen-Hall and $n^{-3/4}$ for the corrected version.
Thus, leaving both bandwidths at their defaults changes smoothing as well
as the correction.

The paper's full $O(n^{-2})$ coverage result requires its stated regularity
conditions and bandwidth restrictions. The limiting factor has a different
accuracy result [@chenhall1993, Section 4]. Neither result automatically
applies to arbitrary data-dependent bandwidths or independent-group
extensions. The package does not implement a separate Bartlett-corrected AEL.

For the paper's simulation kernel, which has support
$[-\sqrt{5},\sqrt{5}]$, use `kernel = "epanechnikov"` and
`bandwidth = sqrt(5) * h_paper`. The package's Epanechnikov option is
parameterized on $[-1,1]$.

## Analyze paired or independent samples

### Paired observations

Rows must identify the same pair in both vectors. The package forms the
differences before sorting and applies the one-sample engine to them:

```{r paired}
set.seed(2702)
before <- rnorm(60)
after <- before + rnorm(60, mean = 0.25, sd = 0.8)

elq_two_sample(
  before, after, paired = TRUE, delta = 0, tau = 0.5,
  method = "chen_hall_bartlett", conf.int = TRUE
)
```

This tests whether the median of `before - after` is zero. Here the
population median difference is $-0.25$. Dependence within each pair is
allowed; independence across pairs is assumed.

### Independent samples

For $q_{\tau,A}-q_{\tau,B}=\delta_0$, the constrained statistic is
\[
T(\delta_0)=
\inf_t\{W_A(t+\delta_0)+W_B(t)\}
-\inf_u W_A(u)-\inf_v W_B(v).
\]
The nuisance parameter $t$ is profiled out. Subtraction of the separate
minima matters for stepwise methods, whose unrestricted minima need not
equal zero. The reference distribution has one degree of freedom.

```{r independent}
set.seed(2703)
A <- rnorm(60)
B <- rnorm(90, mean = 0.4, sd = 2)
two <- elq_two_sample(A, B, delta = 0, tau = 0.5, method = "chen_hall")
two
two$optimizer
two$refinement.gap
```

For independent groups, `"chen_hall_bartlett"` corrects each group's
criterion before profiling. This is an experimental finite-sample
modification with first-order calibration; a joint two-sample
Bartlett-correction theorem has not been established for this implementation.

Independent-difference intervals for all methods are available by subtracting
two marginal confidence hulls. For overall level $1-\alpha$, each marginal
interval uses level $1-\alpha/2$:

```{r independent-intervals}
elq_confint(
  A, B, tau = 0.5, method = "chen_hall", interval = "bonferroni"
)
```

If the marginal intervals are $[L_A,U_A]$ and $[L_B,U_B]$, the returned
interval is $[L_A-U_B,U_A-L_B]$. Its coverage depends on the marginal
calibrations; there is no automatic finite-sample guarantee. It need not
invert the profile test. Direct profile-difference hulls are currently
implemented for `"el"` and `"ael"`:

```{r step-profile-interval}
elq_confint(A, B, method = "el", interval = "profile")
```

## Compare several groups and report post-hoc results

The omnibus null states that all group quantiles at a specified $\tau$
are equal. The package compares a common-quantile fit with separate group
fits:
\[
T=\inf_t\sum_{j=1}^k W_j(t)
-\sum_{j=1}^k\inf_{q_j}W_j(q_j).
\]
Its first-order reference distribution is $\chi^2_{k-1}$, with a fixed
number of independent groups and appropriate regularity conditions.

```{r groups}
set.seed(2704)
groups <- list(A = A, B = B, C = rnorm(75, mean = 0.7, sd = 1.5))
elq_k_sample(groups, tau = 0.5, method = "el")
```

For a data-frame interface, `elq_anova(value ~ treatment, data = data)`
accepts one response and one grouping variable. This is a quantile equality
test; factorial effects and repeated-measures models are not fitted.

For post-hoc analysis, specify the full comparison family in one call:

```{r posthoc}
post <- elq_posthoc(
  groups, tau = c(0.5, 0.75), method = "el",
  p.adjust.method = "holm", conf.int = TRUE
)
knitr::kable(
  post[c("group1", "group2", "tau", "estimate", "p.adj",
         "conf.low", "conf.high")],
  digits = 4, row.names = FALSE
)
```

These three pairs at two quantile levels form **one family of six tests**.
Holm adjustment is applied across all six p-values. The reported intervals
use a separate Bonferroni construction; they are not inversions of
Holm-adjusted p-values. Post-hoc testing is not automatically gated by an
omnibus rejection. The procedure inherits the calibration limits of its
constituent tests.

## Plan power and sample size

### Analytical planning

Analytical power requires the alternative population quantiles $q_j$
and densities $f_j(q_j)$, together with sample sizes $n_j$. A quantile
difference alone does not specify sampling variability:
\[
V_j=\frac{\tau(1-\tau)}{n_j f_j(q_j)^2}.
\]
For a full-row-rank contrast matrix $C$, let $d=Cq-d_0$. The calculator
uses
\[
\lambda=d^\top\{C\operatorname{diag}(V_j)C^\top\}^{-1}d,
\qquad
\Pr\{\chi^2_r(\lambda)>\chi^2_{r,1-\alpha}\},
\]
where $r$ is the number of contrasts. By default, $C$ compares all
groups to the final group, giving $r=k-1$. One sample uses $r=1$.

This is a shared local asymptotic approximation, not method-specific
finite-sample EL power. For a single contrast, directional planning is
also available through `alternative`.

For normal groups with medians 0 and 0.5 and standard deviations 1 and 2:

```{r analytical-power}
q_alt <- c(0, 0.5)
f_alt <- dnorm(0) / c(1, 2)

elq_power(
  n = c(100, 150), quantiles = q_alt, density = f_alt, tau = 0.5
)

plan <- elq_sample_size(
  power = 0.8, quantiles = q_alt, density = f_alt, ratio = c(1, 2)
)
plan
plan$previous.power
```

The search finds the smallest integer base size on the specified allocation
path, with group sizes obtained by rounding up. It does not optimize the
allocation ratio. The previous base size's power helps check the boundary.

```{r power-curve, fig.cap="Analytical omnibus power for two normal populations with medians 0 and 0.5, standard deviations 1 and 2, and allocation n2 = 2 n1. The dashed line marks the requested 80% power."}
base_n <- seq(20, 240, by = 20)
power_values <- vapply(
  base_n,
  function(nn) elq_power(
    n = c(nn, 2 * nn), quantiles = q_alt, density = f_alt
  )$power,
  numeric(1)
)
plot(
  base_n, power_values, type = "b", pch = 19, ylim = c(0, 1),
  xlab = "Sample size in group 1 (group 2 is twice as large)",
  ylab = "Approximate power"
)
abline(h = 0.8, lty = 2, col = "firebrick")
points(plan$base.n, plan$power, pch = 4, cex = 1.4)
```

For paired planning, supply the quantile and density of the **paired
differences** as a one-sample design. The returned sample size is the number
of pairs. For example, normally distributed differences with median 0.4
and standard deviation 1 have density `dnorm(0)` at their median.

`elq_density(x, tau)` supplies a pilot KDE estimate of a density at a
quantile. Its uncertainty is not propagated automatically; inspect plausible
density values and confirm the design by simulation.

### Exact count-based planning

For one-sample ordinary EL, AEL, and sign tests, the statistic depends on
the count $M=\sum_i I(X_i\leq q_0)$. Under a specified alternative,
$M\sim\operatorname{Binomial}(n,p)$, where $p=F_{\mathrm{alt}}(q_0)$.
The calculator enumerates all counts and sums their probabilities over
the rejection region.

```{r exact-power}
# X ~ N(0.5, 1), testing the null median q0 = 0.
p_alt <- pnorm(0, mean = 0.5)
exact_methods <- c("el", "ael", "sign")
exact_results <- lapply(
  exact_methods,
  function(m) elq_power_exact(n = 40, p = p_alt, method = m)
)
knitr::kable(
  data.frame(
    method = exact_methods,
    power = vapply(exact_results, function(z) z$power, numeric(1)),
    null.size = vapply(exact_results, function(z) z$null.size, numeric(1))
  ),
  digits = 4
)

elq_sample_size_exact(
  power = 0.8, p = p_alt, method = "sign", max.n = 300,
  max.null.size = 0.05
)
```

The argument `p` is **not** the alternative quantile or the target level
`tau`. Exact enumeration calculates the rejection probability of the
specified rule; it does not turn a chi-square-calibrated EL rule into an
exact-level test. Inspect `null.size`. Sample-size enumeration checks
integers in ascending order because discrete power need not increase
monotonically. These exact calculators are not implemented for independent
two-sample or k-sample profile tests.

The three methods have the same rejection counts in the example above, so
their enumerated powers agree. Agreement at one setting does not imply that
their finite-sample behavior agrees at other sizes or quantile levels.

### Method-specific simulation

Supply functions that generate each population. The simulation calls the
actual test, so it captures finite-sample features of that method and its
numerical search.

```{r simulated-power}
generators <- list(
  function(n) rnorm(n),
  function(n) rnorm(n, mean = 0.5, sd = 2)
)

sim <- elq_power_sim(
  n = c(40, 60), generators = generators, method = "chen_hall",
  nsim = 50, seed = 2705
)
knitr::kable(
  data.frame(
    power = sim$power, mcse = sim$mcse,
    lower = unname(sim$mc.interval[1]),
    upper = unname(sim$mc.interval[2]),
    successful = sim$valid, failed = sim$failures
  ),
  digits = 4
)
sim$failure.bounds
```

Fifty replicates keep this tutorial quick; they are insufficient for a
precise performance comparison. The Monte Carlo standard error is
$\{\widehat\pi(1-\widehat\pi)/B_{\mathrm{valid}}\}^{1/2}$, and the
reported Monte Carlo interval is a Wilson interval. Power is conditional
on successful replicates; failures, warnings, and bounds obtained by counting
all failures as rejections or nonrejections remain available.

Under a null-generating distribution the same function estimates **Type I
error**, not power against an alternative. Assess that as well as power.
Simulation assessment does not automatically recalibrate the test. These
functions target the omnibus test, not power for Holm-adjusted post-hoc
comparisons.

For a larger study, increase `nsim` and keep the raw replicates. This
deliberately unevaluated example is not run when building the vignette:

```{r research-simulation, eval=FALSE}
research_run <- elq_power_sim(
  n = c(40, 60), generators = generators, method = "chen_hall",
  nsim = 10000, seed = 92705, keep = TRUE
)
research_run$replicates
```

Use the same seed and generators to compare deterministic methods on the
same datasets. A supplied seed restores the caller's RNG state after the
simulation. Paired generators must return an $n$-by-2 matrix with rows
representing pairs.

For a sample-size grid:

```{r simulated-sample-size}
grid_plan <- elq_sample_size_sim(
  n.grid = c(60, 120, 240), generators = generators,
  ratio = c(1, 2), power = 0.8, method = "el",
  nsim = 100, seed = 2706
)
knitr::kable(grid_plan$table, digits = 4)
grid_plan$selected.n
```

Selection is restricted to these candidate sizes. It is not proof of the
globally smallest sample size or a guarantee of achieved population power.
`criterion = "lower"` instead checks the pointwise Monte Carlo lower limit;
it does not create simultaneous or post-selection confidence guarantees.
Confirm a selected design using an independent, larger run.

## Understand the numerical diagnostics

### The EL dual solver

For fixed estimating values $g_i$, `elq_dual()` solves
\[
\sum_i\frac{w_i g_i}{1+\lambda g_i}=0,
\qquad 1+\lambda g_i>0,
\]
using a bracketed scalar root search through `stats::uniroot()`. This is
R's Brent-based root finder, not a generic optimization over all observation
probabilities. Frequencies `weights` compress repeated estimating values;
they are not a general survey-weight inference interface.

```{r dual-example}
dual <- elq_dual(g = c(-0.5, 0.5), weights = c(12, 18))
dual[c("statistic", "lambda", "probabilities", "residual")]
```

The two returned probabilities are aggregate masses for the two estimating
values. They sum to one. The residual measures the estimating constraint's
numerical error. An infinite statistic can represent no positive-likelihood
solution rather than an optimizer failure. If some estimating values are
zero and all others have one sign, a boundary probability vector may satisfy
the moment constraint while still giving likelihood zero.

### Knots and profiling

"Knots" collect breakpoints and additional useful search points. Ordinary
EL/AEL use observed values, where the empirical-CDF count changes. Adimari
uses interpolation breakpoints. Compact-kernel methods include
$X_i-h,X_i,X_i+h$; Zhou-Jing also includes roots of $F_h(q)=\tau$.
For Gaussian smoothing, the $X_i\pm10h$ points are numerical search
points, not finite-support breakpoints.

The independent-group profiler enumerates the stepwise cells exactly.
Adimari searches all interpolation segments. Other smooth criteria use knots,
a grid, local `stats::optimize()` searches, and midpoint refinement.
`optimize()` combines golden-section search and parabolic interpolation.
A small `refinement.gap` shows stability under the implemented refinement,
not a guaranteed global minimum.

Increase the grid resolution to assess numerical sensitivity:

```{r profile-diagnostic}
two_finer <- elq_two_sample(
  A, B, method = "chen_hall",
  control = elq_control(grid.points = 161, refine = TRUE)
)
knitr::kable(
  data.frame(
    grid.points = c(81, 161),
    statistic = c(unname(two$statistic), unname(two_finer$statistic)),
    p.value = c(two$p.value, two_finer$p.value),
    refinement.gap = c(two$refinement.gap, two_finer$refinement.gap)
  ),
  digits = 6
)
```

The sparse-tail warning checks whether
$\min_j n_j\min(\tau,1-\tau)<5$. Five is a heuristic, not a theorem or a
calibration guarantee. Passing this check does not establish accurate Type I
error. `elq_control(warn = FALSE)` suppresses notices; it does not improve
statistical or numerical accuracy.

## Reproduce and extend this workflow

The examples use fixed seeds, local simulated data, and public package
functions. Building the vignette requires no data downloads or user
credentials. The HTML uses native MathML and embedded figures, so it needs
no external mathematics service.

After installing a built package, open the tutorial with:

```{r open-vignette, eval=FALSE}
vignette("quantile-inference", package = "elquantile")
```

Developers can build the package from its parent directory:

```sh
R CMD build elquantile
R CMD check --no-manual elquantile_0.1.0.tar.gz
```

Install `knitr` and `rmarkdown` and make Pandoc available before building
from source. They are vignette-building tools, not dependencies of the
inference functions. Installing the built source archive includes the
rendered vignette; installing the raw source directory alone does not
necessarily build it.

For function arguments, see `?elq_test`, `?elq_confint`, `?elq_power`, and
`?elq_power_sim`. The installed files `doc/METHODS.md` and
`doc/EXTENDING.md` describe the equations and internal engine contract.
New group corrections, clustered-data methods, and general contrast tests
require their own definitions and validation. This tutorial does not
establish their theory or methodological novelty.

The environment used for this rendering is recorded below.

```{r session-information}
sessionInfo()
```

## References
